\systemEnding: Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-06-08 nampi.tex
added unformatted regex figure
shortened

\newcommand{\system}{\textsc{DreamCoder}~}
\newcommand{\systemEnding}{\textsc{DreamCoder}}
\newcommand{\lowerBound}{\mathscr{L}}
\newcommand{\code}[1]{{\footnotesize\texttt{#1}}}
\newcommand{\codechar}[1]{{\footnotesize{\texttt{"#1"}}}}
% Use the following line for the initial blind version submitted for review:
\usepackage[nohyperref]{icml2018}

\usepackage{xcolor}
\definecolor{pop1}{HTML}{1F78b4}
\definecolor{pop2}{HTML}{297F23}
\definecolor{pop3}{HTML}{d95F02}
\definecolor{orange}{HTML}{d95F02}
\definecolor{teal}{HTML}{1b9e77}
\newcommand{\pop}[1]{\textcolor{pop1}{#1}}
\newcommand{\popp}[1]{\textcolor{pop2}{#1}}
\newcommand{\tree}[1]{\textcolor{pop3}{#1}}
\newcommand{\orange}[1]{\textcolor{orange}{#1}}
\newcommand{\teal}[1]{\textcolor{teal}{#1}}

@@

\begin{icmlauthorlist}
\icmlauthor{Kevin Ellis}{MIT}
\icmlauthor{Lucas Morales}{MIT}
\icmlauthor{Matthias}{MIT}
\icmlauthor{Armando Solar-Lezama}{MIT}
\icmlauthor{Joshua B. Tenenbaum}{MIT}
\end{icmlauthorlist}

@@

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

@@

\section{Programs that manipulate sequences}\label{sequences}
We apply \system to list processing (Section~\ref{listSection}) and text editing (Section~\ref{textSection}),
using a bidirectional GRU~\cite{cho2014learning} for
the recognition model, and initially providing the system with generic sequence manipulation primitives:
\code{foldr}, \code{unfold}, \code{if}, \code{map}, \code{length},
\code{index}, \code{=}, \code{+}, \code{-}, \code{0}, \code{1}, \code{cons},
\code{car}, \code{cdr}, \code{nil}, and \code{is-nil}.

\subsection{List Processing}\label{listSection}
Synthesizing programs that manipulate data structures is a widely studied
problem in the programming languages community~\cite{feser2015synthesizing}.
%with applications to computer aided programming~\cite{solar2008program}.
We consider this problem within the context of learning functions that
manipulate lists, and also perform arithmetic operations upon lists (Table~\ref{listExamples}).
\begin{figure}\centering
  \begin{tabular}{lll}
    \toprule
    Name & Input & Output \\\midrule
    repeat-2 & [7\, 0] & [7\, 0\, 7\, 0] \\
%    drop-3 & [0\, 3\, 8\, 6\, 4] & [6\, 4] \\
    rotate-2 & [8\, 14\, 1\, 9] & [1\, 9\, 8\, 14] \\
%    count-head-in-tail & [1\, 2\, 1\, 1\, 3] & 2 \\
    keep-mod-5 & [5\, 9\, 14\, 6\, 3\, 0] & [5\, 0] \\
    product & [7\, 1\, 6\, 2] & 84 \\
    \bottomrule
  \end{tabular}
  \captionof{table}{Some tasks in our list function domain.}\label{listExamples}
\end{figure}
We created 236 human-interpretable list manipulation tasks, each with 15
input/output examples.
%% Our data set is interesting in three major ways: many of the tasks
%% require complex solutions; the tasks were not generated from some
%% latent DSL, and the agent must learn to solve these complicated
%% problems from only 236 tasks.
%% Our data set assumes arithmetic operations as well as sequence operations,
%% so we additionally provide our system with the following arithmetic
%% primitives: \code{mod}, \code{*}, \code{>}, \code{is-square},
%% \code{is-prime}.
%% evaluating  on a 50/50 test/train split.
In solving these tasks, the system
composed 38 new subroutines, and rediscovered the higher-order
function \code{filter} (Tbl.~\ref{initialExampleDSL}, left).%% , yielding a more expressive DSL more closely
%% matching the domain (Tbl.~\ref{initialExampleDSL}, left).

\subsection{Text Editing}\label{textSection}
Synthesizing programs that edit text is a classic problem in the
programming languages and AI literatures~\cite{lau2001programming},
and algorithms that learn text editing programs ship in Microsoft Excel~\cite{gulwani2011automating}.
This prior work presumes a hand-engineered DSL.
Here we will instead start out with generic sequence manipulation
primitives.%%  and recover many of the higher-level building blocks that have
%% made these other text editing systems successful.
%% Because our enumerative search procedure cannot generate string %
%% constants, we instead enumerate programs with string-valued
%% parameters.  For example, to learn a program that prepends ``Dr.'', we
%% enumerate $\text{\code{(}}f_3\code{ string s)}$ -- where $f_3$ is the
%% learned appending primitive (Fig.~\ref{initialExampleDSL}) --- and then
%% define $\probability[x|p]$ by approximately marginalizing out the
%% string parameters via a simple dynamic program.
%% In Sec.~\ref{regressionSection}, we will use a similar trick to
%% synthesize programs containing real numbers, but using gradient
%% descent instead of dynamic programming.

We trained our system on 109 synthetic tasks, with 4 input/output examples each.
After three iterations, it assembles a DSL containing a dozen new functions (Fig.~\ref{initialExampleDSL}, center) solving
all the training tasks.
But, how well does the  learned DSL generalized to real text-editing scenarios?
We tested, but did not train, on the 108 text editing problems from the SyGuS~\cite{alur2016sygus} program synthesis competition. Before any learning,
\system solves 3.7\% of the problems with an average search time of 235 seconds.
After learning,
it solves 74.1\%, and does so much faster,
solving them in an average of 29 seconds.
As of the 2017 SyGuS competition,
the best-performing algorithm solves 79.6\% of the problems.
But, SyGuS comes with a
different hand-engineered DSL \emph{for each text editing problem}.%\footnote{SyGuS text editing problems also prespecify the set of allowed string constants for each task. For these experiments, our system did not use this assistance.}
Here  we learned a single DSL
that applied generically to
all of the tasks,
and perform comparably to the best
prior work.

\section{Symbolic Regression: Programs from visual input}\label{regressionSection}
We apply \system
to symbolic regression problems.  Here, the
agent observes points along the curve of a function, and must write a
program that fits those points.  We initially equip our learner with
addition, multiplication, and division, and task it with solving
100 symbolic regression problems, each either a polynomial or rational function.  The recognition model is a
convnet that observes an image of the target function's
graph (Fig.~\ref{functions}) --- visually, different kinds of
polynomials and rational functions produce different kinds of graphs,
and so the recognition model can  look at a graph and predict
what kind of function best explains it.
These programs can contain real numbers,
which we set with gradient descent,
and penalize continuous parameters by
incorporating a BIC penalty into the likelihood model $\probability[x|p]$.
%% A key difficulty, however, is
%% that these problems are best solved with programs containing real
%% numbers.  Our solution to this difficulty is to enumerate
%%  programs with real-valued parameters, and then fit those
%% parameters by automatically differentiating through the programs the
%% system writes and use gradient descent to fit the parameters.
%% We define the likelihood model, $\probability[x|p]$, by assuming a Gaussian noise model for the input/output examples,
%% and penalize the use of real-valued parameters using the BIC~\cite{Bishop:2006:PRM:1162264}.
\begin{figure}\vspace{-0.0cm} \newcommand{\functionSize}{1.1cm}\centering
  \begin{minipage}[c]{0.3\columnwidth}
  \includegraphics[width = \functionSize]{figures/functions/6.png}
  \includegraphics[width = \functionSize]{figures/functions/48.png}\\
  \includegraphics[width = \functionSize]{figures/functions/102.png}
  \includegraphics[width = \functionSize]{figures/functions/116.png}\\
  \vspace{1pt}%\includegraphics[width = \functionSize]{figures/functions/181.png}
  %\includegraphics[width = \functionSize]{figures/functions/160.png}
  \includegraphics[width = \functionSize]{figures/functions/160.png}
    \includegraphics[width = \functionSize]{figures/functions/149.png}
  \end{minipage}
  \begin{minipage}[c]{0.69\columnwidth}    
    \caption{Recognition model input for symbolic regression. DSL learns subroutines for polynomials (top rows) and rational functions (bottom) while the recognition  model jointly learns to look at a graph of the function (left) and predict which learned subroutines best explain the observation.}\label{functions}
        \end{minipage}
\end{figure}
We learn a DSL containing 13 new functions,
mainly templates for polynomials of different orders or ratios of polynomials.
It also learns to find programs minimizing the number of continuous parameters ---
learning to represent linear functions with 
\code{(* real (+ x real))}, which has two continuous parameters, and represents quartic functions using $f_4$ in the rightmost column of Fig.~\ref{initialExampleDSL}
which has five continuous parameters.
This phenomenon arises from our Bayesian framing.

\section{Learning from Scratch}
A long-standing dream within the program induction community
is ``learning from scratch'': starting with a \emph{minimal} Turing-complete programming language,
and then learning to solve a wide swath of
induction problems~\cite{solomonoff1964formal}.
All existing systems,
including ours,
fall far short of this dream,
and it is unclear (and we believe unlikely)
that this dream could ever be fully realized.
How far can we push in this direction?
``Learning from scratch'' is subjective, but a reasonable
starting point is the set of primitives provided in 1959
Lisp~\cite{mccarthy1960recursive}: these include
conditionals, recursion, arithmetic, and the 
list operators \code{cons}, \code{car}, \code{cdr}, and \code{nil}.
A  basic first goal is to start with
these primitives,
and then recover a DSL that
more closely resembles modern functional languages like Haskell and OCaml.
Recall (Sec.~\ref{sequences})
that we initially provided our system with functional programming routines like
\code{map} and \code{fold}.

@@

\section{Learning Generative Models}

We apply \system to learning generative models for images (Figure~\ref{geomCompiled}) and text (Figure~\ref{regularExpressions}).
For images, we learn programs in a LOGO/Turtle-like representation,
and the task is to look at an image and explain it in terms of a graphics program.
For text, we learn probabilistic regular expressions -- a simple probabilistic program for which inference is always tractable -- and the task is to
infer a regex from a set of strings.
\begin{figure}[t]
  \includegraphics[width=\columnwidth]{figures/geomCompiled.eps} 
  \caption{
  Top left: Example training tasks. Top right: samples from the learned DSL.
 Bottom: \% holdout testing tasks solved (middle); on the sides are averaged samples from the DSL
  before any training (left) and after last iteration (right).}\label{geomCompiled}

  \end{figure}

\section{Quantitative Results}\label{quantitative}
We evaluate  on held-out testing tasks,
measuring how many
tasks are solved and how long it takes to solve them (Fig.~\ref{learningCurves}).
Prior to any learning,
the system cannot find solutions for most of the tasks,
and those it does solve take a long time;
with more wake/sleep iterations,
we converge upon DSLs and recognition models 
more closely matching the domain.

>>>>

\newcommand{\system}{\textsc{DreamCoder}~}
\newcommand{\systemEnding}{\textsc{DreamCoder}}
\newcommand{\lowerBound}{\mathscr{L}}
\newcommand{\code}[1]{{\footnotesize\texttt{#1}}}
\newcommand{\scode}[1]{{\tiny\texttt{#1}}}
\newcommand{\codechar}[1]{{\footnotesize{\texttt{"#1"}}}}
% Use the following line for the initial blind version submitted for review:
\usepackage[nohyperref]{icml2018}

\usepackage{xcolor}
\definecolor{pop1}{HTML}{1F78b4}
\definecolor{pop2}{HTML}{164C13}
\definecolor{pop3}{HTML}{d95F02}
\definecolor{orange}{HTML}{d95F02}
\definecolor{teal}{HTML}{1b9e77}
\newcommand{\pop}[1]{\textcolor{pop1}{#1}}
\newcommand{\popp}[1]{\textcolor{pop2}{#1}}
\newcommand{\tree}[1]{\textcolor{pop3}{#1}}
\newcommand{\orange}[1]{\textcolor{orange}{#1}}
\newcommand{\teal}[1]{\textcolor{teal}{#1}}

@@

\begin{icmlauthorlist}
\icmlauthor{Kevin Ellis}{MIT}
\icmlauthor{Lucas Morales}{MIT}
\icmlauthor{Matthias}{MIT}
\icmlauthor{Maxwell Nye}{MIT}
\icmlauthor{Armando Solar-Lezama}{MIT}
\icmlauthor{Joshua B. Tenenbaum}{MIT}
\end{icmlauthorlist}

@@

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

@@

\section{Experiments}

\subsection{Programs that manipulate sequences}\label{sequences}
We apply \system to list processing %% (Section~\ref{listSection})
and text editing, %% (Section~\ref{textSection})
using a bidirectional GRU~\cite{cho2014learning} for
the recognition model, and initially providing the system with generic sequence manipulation primitives:
\code{foldr}, \code{unfold}, \code{if}, \code{map}, \code{length},
\code{index}, \code{=}, \code{+}, \code{-}, \code{0}, \code{1}, \code{cons},
\code{car}, \code{cdr}, \code{nil}, and \code{is-nil}.

%\subsubsection{List Processing}\label{listSection}
\textbf{List Processing:} Synthesizing programs that manipulate data structures is a widely studied
problem in the programming languages community~\cite{feser2015synthesizing}.
%with applications to computer aided programming~\cite{solar2008program}.
We consider this problem within the context of learning functions that
manipulate lists, and also perform arithmetic operations upon lists (Table~\ref{listExamples}).
We created 236 human-interpretable list manipulation tasks, each with 15
input/output examples.
%% Our data set is interesting in three major ways: many of the tasks
%% require complex solutions; the tasks were not generated from some
%% latent DSL, and the agent must learn to solve these complicated
%% problems from only 236 tasks.
%% Our data set assumes arithmetic operations as well as sequence operations,
%% so we additionally provide our system with the following arithmetic
%% primitives: \code{mod}, \code{*}, \code{>}, \code{is-square},
%% \code{is-prime}.
%% evaluating  on a 50/50 test/train split.
In solving these tasks, the system
composed 38 new subroutines, and rediscovered the higher-order
function \code{filter} ($f_1$ in Table~\ref{initialExampleDSL}, left).%% , yielding a more expressive DSL more closely
%% matching the domain (Tbl.~\ref{initialExampleDSL}, left).
\begin{figure}[b]\centering
  \begin{tabular}{lll}
    \toprule
    Name & Input & Output \\\midrule
    repeat-2 & [7\, 0] & [7\, 0\, 7\, 0] \\
%    drop-3 & [0\, 3\, 8\, 6\, 4] & [6\, 4] \\
    rotate-2 & [8\, 14\, 1\, 9] & [1\, 9\, 8\, 14] \\
%    count-head-in-tail & [1\, 2\, 1\, 1\, 3] & 2 \\
    keep-mod-5 & [5\, 9\, 14\, 6\, 3\, 0] & [5\, 0] \\
    product & [7\, 1\, 6\, 2] & 84 \\
    \bottomrule
  \end{tabular}
  \captionof{table}{Some tasks in our list function domain.}\label{listExamples}
\end{figure}

%\subsubsection{Text Editing}\label{textSection}
\textbf{Text Editing:} Synthesizing programs that edit text is a classic problem in the
programming languages and AI literatures~\cite{gulwani2011automating,lau2001programming}.
This prior work uses hand-engineered DSLs.
Here, we instead start out with generic sequence manipulation
primitives and recover many of the higher-level building blocks that 
have made these other systems successful.
%% Because our enumerative search procedure cannot generate string %
%% constants, we instead enumerate programs with string-valued
%% parameters.  For example, to learn a program that prepends ``Dr.'', we
%% enumerate $\text{\code{(}}f_3\code{ string s)}$ -- where $f_3$ is the
%% learned appending primitive (Fig.~\ref{initialExampleDSL}) --- and then
%% define $\probability[x|p]$ by approximately marginalizing out the
%% string parameters via a simple dynamic program.
%% In Sec.~\ref{regressionSection}, we will use a similar trick to
%% synthesize programs containing real numbers, but using gradient
%% descent instead of dynamic programming.

We trained our system on 109 automatically-generated text editing tasks, with 4 input/output examples each.
After three iterations, it assembles a DSL containing a dozen new functions (Fig.~\ref{initialExampleDSL}, center) solving
all the training tasks.
But, how well does the  learned DSL generalized to real text-editing scenarios?
We tested, but did not train, on the 108 text editing problems from the SyGuS~\cite{alur2016sygus} program synthesis competition. Before any learning,
\system solves 3.7\% of the problems with an average search time of 235 seconds.
After learning,
it solves 74.1\%, and does so much faster,
solving them in an average of 29 seconds.
As of the 2017 SyGuS competition,
the best-performing algorithm solves 79.6\% of the problems.
But, SyGuS comes with a
different hand-engineered DSL \emph{for each text editing problem}. %\footnote{SyGuS text editing problems also prespecify the set of allowed string constants for each task. For these experiments, our system did not use this assistance.}
Here  we learned a single DSL
that applied generically to
all of the tasks,
and perform comparably to the best
prior work.

\subsection{Symbolic Regression: Programs from visual input}\label{regressionSection}
We apply \system
to symbolic regression problems.  Here, the
agent observes points along the curve of a function, and must write a
program that fits those points.  We initially equip our learner with
addition, multiplication, and division, and task it with solving
100 symbolic regression problems, each either a polynomial or rational function.  The recognition model is a
convnet that observes an image of the target function's
graph (Fig.~\ref{functions}) --- visually, different kinds of
polynomials and rational functions produce different kinds of graphs,
and so the recognition model can  look at a graph and predict
what kind of function best explains it.
These programs can contain real numbers,
which we set with gradient descent,
and penalize continuous parameters by
incorporating a BIC penalty into the likelihood model $\probability[x|p]$.
%% A key difficulty, however, is
%% that these problems are best solved with programs containing real
%% numbers.  Our solution to this difficulty is to enumerate
%%  programs with real-valued parameters, and then fit those
%% parameters by automatically differentiating through the programs the
%% system writes and use gradient descent to fit the parameters.
%% We define the likelihood model, $\probability[x|p]$, by assuming a Gaussian noise model for the input/output examples,
%% and penalize the use of real-valued parameters using the BIC~\cite{Bishop:2006:PRM:1162264}.
\begin{figure}\vspace{-0.0cm} \newcommand{\functionSize}{1.1cm}\centering
  \begin{minipage}[c]{0.3\columnwidth}
  \includegraphics[width = \functionSize]{figures/functions/6.png}
  \includegraphics[width = \functionSize]{figures/functions/48.png}\\
  \includegraphics[width = \functionSize]{figures/functions/102.png}
  \includegraphics[width = \functionSize]{figures/functions/116.png}\\
  \vspace{1pt}%\includegraphics[width = \functionSize]{figures/functions/181.png}
  %\includegraphics[width = \functionSize]{figures/functions/160.png}
  \includegraphics[width = \functionSize]{figures/functions/160.png}
    \includegraphics[width = \functionSize]{figures/functions/149.png}
  \end{minipage}
  \begin{minipage}[c]{0.69\columnwidth}    
    \caption{Recognition model input for symbolic regression. DSL learns subroutines for polynomials (top rows) and rational functions (bottom) while the recognition  model jointly learns to look at a graph of the function (left) and predict which learned subroutines best explain the observation.}\label{functions}
        \end{minipage}
\end{figure}
We learn a DSL containing 13 new functions,
mainly templates for polynomials of different orders or ratios of polynomials.
The model also learns to find programs minimizing the number of continuous parameters ---
learning to represent linear functions with 
\code{(* real (+ x real))}, which has two continuous parameters, and represents quartic functions using $f_4$ in the rightmost column of Fig.~\ref{initialExampleDSL}
which has five continuous parameters.
This phenomenon arises from our Bayesian framing.

\subsection{Learning from Scratch}
A long-standing dream within the program induction community
is ``learning from scratch'': starting with a \emph{minimal} Turing-complete programming language,
and then learning to solve a wide swath of
induction problems~\cite{solomonoff1964formal}.
All existing systems,
including ours,
fall far short of this dream,
%and it is unclear (and we believe unlikely)
and we believe it is unlikely
that this dream could ever be fully realized.
%How far can we push in this direction?
``Learning from scratch'' is subjective, but a reasonable
starting point is the set of primitives provided in 1959
Lisp~\cite{mccarthy1960recursive}: these include
conditionals, recursion, arithmetic, and the 
list operators \code{cons}, \code{car}, \code{cdr}, and \code{nil}.
A  basic first goal is to start with
these primitives,
and then recover a DSL that
more closely resembles modern functional languages like Haskell and OCaml.
Recall (Sec.~\ref{sequences})
that we initially provided our system with functional programming routines like
\code{map} and \code{fold}.

@@

\subsection{Learning Generative Models}

We apply \system to learning generative models for images and text (Fig.~\ref{geomCompiled}-\ref{regularExpressions}).
For images, we learn programs controlling a simulated ``pen,''
and the task is to look at an image and explain it in terms of a graphics program.
For text, we learn probabilistic regular expressions -- a simple probabilistic program for which inference is always tractable -- and the task is to
infer a regex from a collection of strings.
\begin{figure}[t]
  \includegraphics[width=\columnwidth]{figures/geomCompiled.eps} 
  \caption{
  Top left: Example training tasks. Top right: samples from the learned DSL.
 Bottom: \% holdout testing tasks solved (middle); on the sides are averaged samples from the DSL
  before any training (left) and after last iteration (right).}\label{geomCompiled}

  \end{figure}
\begin{figure}\centering
\tiny
  \begin{tabular}{lll}
\multicolumn{3}{c}{Tasks and synthesized programs:}\\
cut&F &Moss Side \\ 
control&CL &Burnage \\
  control&F & City Centre\\
cut& PCFL& Brooklands \\
    \hline
\scode{\textbackslash l*\textbackslash l} & \scode{((\textbackslash u\textbackslash u)*)\textbar F}& \scode{( \textbar\textbackslash u\textbackslash w*)*} \\
    \bottomrule
  \end{tabular}
  \quad
    \begin{tabular}{l}
\multicolumn{1}{c}{DSL:}\\
$f_1 = $ \scode{\textbackslash u\textbackslash w*}\\
$f_2(x) = $ \scode{($x$|$f_1$)*} $=$ \scode{($x$|\textbackslash u\textbackslash w*)*}\\
$f_3(x) =  f_2(\scode{space}) = $\scode{( \textbar\textbackslash u\textbackslash w*)*}\\
\hline
$f_4(x) = $ \scode{($x$*$x$)} (equivalent to regex `plus')\\
$f_5 = f_4(\scode{\textbackslash l}) = \scode{\textbackslash l*\textbackslash l}$\\
    \bottomrule
  \end{tabular}
\\
\begin{tabular}{ccc}
\multicolumn{3}{c}{samples from synthesized programs:}\\
 \scode{\textbackslash l*\textbackslash l} & \scode{((\textbackslash u\textbackslash u)*)\textbar F}& \scode{( \textbar\textbackslash u\textbackslash w*)*} \\
\hline
ya&DQDF & Vr DR \\ 
fcbadcvoaobvjf&F &BeF lKQ \\
mgs&F& W\\
piljnl& KI& kqBfZ 0 \\
    \bottomrule
  \end{tabular}
            \centering 
            \includegraphics[width=.25\textwidth]{figures/nampiregexjoshcurve.png}   
\caption{regex stuff}\label{regularExpressions}
\end{figure}

\subsection{Quantitative Results}\label{quantitative}
We evaluate  on held-out testing tasks,
measuring how many
tasks are solved and how long it takes to solve them (Fig.~\ref{learningCurves}).
Prior to any learning,
the system cannot find solutions for most of the tasks,
and those it does solve take a long time;
with more wake/sleep iterations,
we converge upon DSLs and recognition models 
more closely matching the domain.
